---
title: "pre_processing_metabolomics"
author: "TP"
date: "2024-11-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 0. Packages
```{r}
library(ggplot2)
library(dplyr)
library(UpSetR)
library(naniar)
```


# 1. Load data
## ** metabolomics
```{r}
metabo <- read.delim("./240323_metabolomics_ionmatrix_withannotation_avg.txt", header = T)
rownames(metabo) <- metabo$ionIdx
head(metabo)
```

```{r}
metabo_mtx <- metabo[,c(7:ncol(metabo))]
rownames(metabo_mtx) <- metabo$ionMz
rownames(metabo_mtx)[1:10]
all(sapply(metabo_mtx, is.numeric))
```
```{r}
head(metabo_mtx)
```

```{r}
dim(metabo_mtx)
```

```{r}
metabo_anno <- metabo[,c(1:6)]
metabo_anno$ion_name <- paste0("ion_",metabo_anno$ionIdx)
head(metabo_anno)
```




## ion annotation
```{r}
ion_anno <- read.delim("./240324_ion_anno.txt", header = T)

head(ion_anno)
```
### add database
```{r}
ion_anno$database <- "KEGG"
ion_anno[grepl("CHEBI", ion_anno$CompoundID),"database"] <- "ChEBI"
ion_anno[grepl("HMDB", ion_anno$CompoundID),"database"] <- "HMDB"
```

### add source annotation
```{r}
## how many samples are from which database
count_ids_perdatabase <- ion_anno %>%
                          group_by(database) %>%
                          summarize(count.idx = n_distinct(ionMz))

count_ids_perdatabase <- as.data.frame(count_ids_perdatabase)

ggplot(count_ids_perdatabase, aes(x= database, y= count.idx))+
  geom_bar(stat = "identity")+
  geom_text(aes(label = count.idx), color = "white", vjust = 1)+
  theme(aspect.ratio = 1.5)
```
```{r}
list_database <- split(
  ion_anno$ionMz, 
  ion_anno$database
)

UpSetR::upset(fromList(list_database))
```
*--> unique peaks for each database will remain with this database*
*--> peaks that are shared between ChEBI and KEGG or HMDB --> will be annotated by KEGG or HMDB*
*--> the rest - which are shared with HMDB will be annotated by HMDB*

```{r}
database_mtx <- as.data.frame( ComplexHeatmap::list_to_matrix(list_database))

head(database_mtx)
```

```{r}
database_mtx[database_mtx$ChEBI == 1,"ChEBI"] <- "ChEBI"
database_mtx[database_mtx$ChEBI == 0,"ChEBI"] <- ""

database_mtx[database_mtx$HMDB == 1,"HMDB"] <- "HMDB"
database_mtx[database_mtx$HMDB == 0,"HMDB"] <- ""

database_mtx[database_mtx$KEGG == 1,"KEGG"] <- "KEGG"
database_mtx[database_mtx$KEGG == 0,"KEGG"] <- ""

head(database_mtx)

```

```{r}
database_mtx$intersect <- apply(database_mtx,1,paste, collapse = "")
database_mtx$ionMz <- rownames(database_mtx)

head(database_mtx)
```

### Annoate by HMDB
```{r}
annotate_list <- split(
  database_mtx$ionMz,
  database_mtx$intersect
)

str(annotate_list)
```

```{r}
ion_annotation_tal <- list()

## 1. chebi
tmp <- NULL
tmp <- ion_anno[ion_anno$ionMz %in% annotate_list$ChEBI,]

tmp <- tmp %>%
  group_by(ionMz) %>%
  filter(row_number()==1)

tmp$annotation <- "ChEBI"

ion_annotation_tal[["ChEBI"]] <- tmp

# ==========================================

## 2. KEGG
tmp <- NULL
tmp <- ion_anno[ion_anno$ionMz %in% annotate_list$KEGG,]

tmp <- tmp %>%
  group_by(ionMz) %>%
  filter(row_number()==1)

tmp$annotation <- "KEGG"

ion_annotation_tal[["KEGG"]] <- tmp

# ==========================================

## 3. HMDB
tmp <- NULL
tmp <- ion_anno[ion_anno$ionMz %in% annotate_list$HMDB,]

tmp <- tmp %>%
  group_by(ionMz) %>%
  filter(row_number()==1)

tmp$annotation <- "HMDB"

ion_annotation_tal[["HMDB"]] <- tmp

# ==========================================

## 4. ChEBIKEGG
tmp <- NULL
tmp <- ion_anno[ion_anno$ionMz %in% annotate_list$ChEBIKEGG,]

tmp <- tmp %>%
  group_by(ionMz) %>%
  filter(database == "KEGG") %>%
  filter(row_number()==1)

tmp$annotation <- "ChEBIKEGG"

ion_annotation_tal[["ChEBIKEGG"]] <- tmp

# ==========================================

## 5. ChEBIHMDB
tmp <- NULL
tmp <- ion_anno[ion_anno$ionMz %in% annotate_list$ChEBIHMDB,]

tmp <- tmp %>%
  group_by(ionMz) %>%
  filter(database == "HMDB") %>%
  filter(row_number()==1)

tmp$annotation <- "ChEBIHMDB"

ion_annotation_tal[["ChEBIHMDB"]] <- tmp

# ===========================================

## 6. HMDBKEGG
tmp <- NULL
tmp <- ion_anno[ion_anno$ionMz %in% annotate_list$HMDBKEGG,]

tmp <- tmp %>%
  group_by(ionMz) %>%
  filter(database == "HMDB") %>%
  filter(row_number()==1)

tmp$annotation <- "HMDBKEGG"

ion_annotation_tal[["HMDBKEGG"]] <- tmp

# ============================================

## 7. ChEBIHMDBKEGG
tmp <- NULL
tmp <- ion_anno[ion_anno$ionMz %in% annotate_list$ChEBIHMDBKEGG,]

tmp <- tmp %>%
  group_by(ionMz) %>%
  filter(database == "HMDB") %>%
  filter(row_number()==1)

tmp$annotation <- "ChEBIHMDBKEGG"

ion_annotation_tal[["ChEBIHMDBKEGG"]] <- tmp
```

```{r}
melt_annotation_tal <- bind_rows(ion_annotation_tal)
melt_annotation_tal <- as.data.frame(melt_annotation_tal)
rownames(melt_annotation_tal) <- melt_annotation_tal$ionMz
## add ionIndx 
melt_annotation_tal <- melt_annotation_tal[order(melt_annotation_tal$ionMz, decreasing = F),]
melt_annotation_tal$ionInx <- c(1:nrow(melt_annotation_tal))

## as.data.frame
melt_annotation_tal <- as.data.frame(melt_annotation_tal)

nrow(melt_annotation_tal) == length(unique(ion_anno$ionMz))
```

```{r}
ggplot(melt_annotation_tal, aes(x= database))+
  geom_bar(stat = "count")+
  geom_text(vjust =1, color = "white", aes(label = after_stat(count)), stat = "count")
```

### ... use manual converstion to HMDB from metaboanalyst
was done on 27.3.24
* CHEBI annotations had no results
* KEGG annotations were partical
```{r}
kegg_to_hmdb <- read.csv("./240327_keggtohmdb_metaboanalyst.csv")
colnames(kegg_to_hmdb)[1] <- "kegg_id"


```
```{r}
nrow(kegg_to_hmdb[!is.na(kegg_to_hmdb$HMDB),])
melt_annotation_tal$kegg_to_hmdb <- kegg_to_hmdb[match(melt_annotation_tal$CompoundID, kegg_to_hmdb$kegg_id),"HMDB"]
```


which KEGG ids were not found at all? 
these will be removed from the metabolites 
```{r}
kegg_miss <- kegg_to_hmdb[is.na(kegg_to_hmdb$KEGG),"kegg_id"]
length(kegg_miss)
kegg_miss[1:10]
```
```{r}
melt_annotation_tal$kegg_miss <- NA
melt_annotation_tal[which(melt_annotation_tal$CompoundID %in% kegg_miss),"kegg_miss"] <- "missing"
```

since all CHEBI were not found in the database

### ... NEW anno: compound ID kegg to hmdb and exclude CHEBI
```{r}
## exclude chebi
melt_annotation_tal_new <- melt_annotation_tal[which(melt_annotation_tal$database != "ChEBI"),]

melt_annotation_tal_new <- melt_annotation_tal_new[is.na(melt_annotation_tal_new$kegg_miss),]

melt_annotation_tal_new$CompoundID_new <- ifelse(
  melt_annotation_tal_new$database == "HMDB", melt_annotation_tal_new$CompoundID,
  ifelse(melt_annotation_tal_new$database == "KEGG", melt_annotation_tal_new$kegg_to_hmdb,NA)
)

## exclude those with no new CompoundID
melt_annotation_tal_new <- melt_annotation_tal_new[!is.na(melt_annotation_tal_new$CompoundID_new),]

melt_annotation_tal_new$database_new <- ifelse(grepl("HMDB", melt_annotation_tal_new$CompoundID_new), "HMDB", "error")

head(melt_annotation_tal_new)
```

```{r}
ggplot(melt_annotation_tal_new, aes(x= database_new))+
  geom_bar(stat = "count")+
  geom_text(vjust =-1, color = "red", aes(label = after_stat(count)), stat = "count")
```
## New metabo_mtx 
```{r}
metabo_mtx_full <- metabo[,c(7:ncol(metabo))]
rownames(metabo_mtx_full) <- metabo$ionMz
head(metabo_mtx_full)
```

```{r}
metabo_mtx <- metabo_mtx_full[which(rownames(metabo_mtx_full) %in% as.character(melt_annotation_tal_new$ionMz)),]


head(metabo_mtx)
```
```{r}
all(as.character(melt_annotation_tal_new$ionMz) %in% rownames(metabo_mtx))
```



# 2. Injection order
```{r }
injections <- read.delim("./240324_injection_order.txt", header = T)

injections$type <- "sample"

injections[grepl("pSS",injections$ID),"type"] <- "pSS"

injections$y <- ifelse(injections$type == "sample",1,2)

ggplot(injections, aes(x= order, y= y, color= type))+
  geom_point(size = 1)+
  scale_color_manual(values = c("sample" = "black", "pSS" = "red"))+
  ylim(0,5)
```



# 3. Pre processing
https://labplan.ie/content/uploads/2018/01/Biocrates-Tutorial-MetaboAnalyst-V01.pdf
I chose to only perform log2 transformation without pareto scaling to avoid problems with data analysis. 

## Log2 normalization
```{r fig.width=20}
#boxplot metabolites before normalization
boxplot(metabo_mtx)
```


```{r fig.width=15}
#200 metabolites example plot before normalization
boxplot(t(metabo_mtx[1:200,]))
```


```{r}
#log2 normalization of metabolites
metabo_log2<-apply(metabo_mtx,2, function(x){log2(x+1)})
rownames(metabo_log2)<-rownames(metabo_mtx)
```

```{r}
## visualize distribution of samples after log2 transformation
hist(metabo_log2, breaks = 50, freq=FALSE)
curve(dnorm(x,mean=mean(metabo_log2),sd=sd(metabo_log2)), add=TRUE,col="red")
```
```{r}
vis_miss(as.data.frame(metabo_log2))
```




### metabolites boxplot

```{r fig.width=20}
topvar_metabo<- order(Rfast::colVars(metabo_log2), decreasing = T)[1:50]
boxplot(metabo_log2[,topvar_metabo])
```

# 4. Session info
```{r}
sessionInfo()
```

